BSTERGM: Bayesian Separable-Temporal Exponential family Graph Models

### 2020-10-26 presentation_simple/presentation_simple.tex
add data and simple version tex

\section{Model definitions and related terminologies}
\begin{frame}{Random graphs}
    \begin{defn}[Random graphs and related terminologies]
    Let $\mathcal{B}=\{0,1\}$. For given $n\in \mathbb{N}$,
    \begin{itemize}
        \item The set $\mathcal{Y} \subset \mathcal{B}^{n^2}$ is a set of graphs of $n$ nodes (without weights for each nodes and edges.)
        \item Let $\Omega$ be an event set. We say $Y: \Omega \to \mathcal{Y}$ is a random variable for a graph, or a random graph.
        \item For a random graph $Y \in \mathcal{Y}$, denote the edge between i-th node and j-th node by $Y_{ij}$ for $i,j=1,2,...,n$, \\
            satisfying $Y_{ij}=1$ if the edge is connected. Otherwise, $Y_{ij}=0$.
        \item If edges of $Y$ have directions, then $Y$ is called a directed graph. Otherwise, $Y$ is called a undirected graph.
    \end{itemize}
    \end{defn}
    Let me notate a realization of random graph by $y$ and its edges by $y_{ij}$ for $i,j=1,2,...,n$.

    Here is a remark. These are obvious that, for given $n\in \mathbb{N}$,
    \begin{itemize}
        \item $|\mathcal{Y}|=2^{n(n-1)}$ if $\mathcal{Y}$ is the set of all directed graphs not permitting self-connecting edges.
        \item $|\mathcal{Y}|=2^{n(n-1)/2}$ if $\mathcal{Y}$ is the set of all graphs of undirected one.
    \end{itemize}
    ;thus, the size of $\mathcal{Y}$ grows exponentially when n increases.
\end{frame}

\begin{frame}{ERGM: Exponential family Random Graphs Models}
    \begin{defn}[ERGM: Exponential family Random Graphs Models]
        Let $\mathcal{Y}$ be a set of graphs with $n$ nodes and $Y$ be a random graph (on $\mathcal{Y}$.)
        Set a distribution on $\mathcal{Y}$ to
        \[P(Y=y;\theta) = \frac{exp(\theta^{T}s(y))}{c(\theta)}\]
        for some $\theta\in\mathbb{R}^p$,
        where $s(y)\in\mathbb{R}^p$ is a vector which is part of y's sufficient network statistics,
        \\ and $c(\theta)\in\mathbb{R}$ is a normalizing constant satisfying $c(\theta)=\sum_{y\in\mathcal{Y}}exp(\theta^{T}s(y))$.
        \\ Models which have a such form called the ERGM.
    \end{defn}

    \begin{defn}[TERGM: Temporal Exponential family Random Graphs Models]
        Let $\mathcal{Y}$ be a set of graphs with $n$ nodes. 
        Let $Y_1=y_1 \in \mathcal{Y}$ be given and $Y_2,...,Y_T (T\in\mathbb{N})$ be random graphs (on $\mathcal{Y}$).
        Set a distribution on $\mathcal{Y}\times ... \times \mathcal{Y}$ ($T-1$ folds) to
        \[P(Y_t=y_t|Y_{t-1}=y_{t-1};\theta) = \frac{exp(\theta^{T}s(y_t, y_{t-1}))}{c(\theta, y_{t-1})}\]
        for $\theta\in\mathbb{R}^p$ and with the first-order Markov assumption $P(Y_2,Y_3,...,Y_T|Y_1)=P(Y_2|Y_1)P(Y_3|Y_2)...P(Y_T|Y_{T-1})$,
        \\where $s(y_t, y_{t-1})\in\mathbb{R}^n$ is a part of sufficient network statistics,
        and $c(\theta, y_{t-1})=\sum_{y\in\mathcal{Y}}exp(\theta^{T}s(y, y_{t-1}))$ is a normalizing constant.
        Models which have a such form called the TERGM.
    \end{defn}
\end{frame}

\begin{frame}{STERGM: Separable-Temporal Exponential family Random Graphs Models}
    \begin{defn}[STERGM: Separable-Temporal Exponential family Random Graphs Models]
        Let $\mathcal{Y}$ be a set of graphs with $n$ nodes. 
        Let $Y_1=y_1 \in \mathcal{Y}$ be given and $Y_2,...,Y_T (T\in\mathbb{N})$ be random graphs (on $\mathcal{Y}$).
        Set a distribution on $\mathcal{Y}\times ... \times \mathcal{Y}$ ($T-1$ folds) by following way:

        Let $\mathcal{Y}^+|_t$ be a subset of $\mathcal{Y}$ consisting all graphs which have equal or additional edges comparing to $y_{t-1}$.
        Likewise, let $\mathcal{Y}^-|_t$ be a subset of $\mathcal{Y}$ consisting all graphs which have equal or sparse edges comparing to $y_{t-1}$.
        Next, for $Y_t^+: \Omega \to\mathcal{Y}^+|_t$, $Y_t^-: \Omega \to\mathcal{Y}^-|_t$, $y_t^+ \in \mathcal{Y}^+|_t$ and $y_t^- \in \mathcal{Y}^-|_t$, set
        \[P(Y_t^+=y_t^+|Y_{t-1}=y_{t-1};\theta^+) = \frac{exp((\theta^+)^{T}s(y_t^+, y_{t-1}))}{c(\theta^+, y_{t-1})},
        P(Y_t^-=y_t^-|Y_{t-1}=y_{t-1};\theta^-) = \frac{exp((\theta^-)^{T}s(y_t^-, y_{t-1}))}{c(\theta^-, y_{t-1})}\]
        for some $\theta^+,\theta^-\in\mathbb{R}^p$, $s(y_t^+, y_{t-1}), s(y_t^-, y_{t-1})\in\mathbb{R}^n$, which are parts of sufficient network statistics,
        and normalizers $c(\theta^+, y_{t-1})=\sum_{y^+\in\mathcal{Y}^+}exp((\theta^+)^{T}s(y^+, y_{t-1})), c(\theta^-, y_{t-1})=\sum_{y^-\in\mathcal{Y}^-}exp((\theta^-)^{T}s(y^-, y_{t-1}))$.

        Then, defining operations $+,-$ on $\mathcal{Y}$ following the boolean algebra edgewise-ly, set $y_t$ to
        \[y_t=y_t^+ - (y_{t-1} - y_t^-) = y_t^- + (y_t^+ - y_{t-1})\]

        Additionally, assume that
        \begin{itemize}
            \item The first-order Markov assumption: $P(Y_2,...,Y_T|Y_1)=P(Y_2|Y_1)...P(Y_T|Y_{T-1})$
            \item The separability: the conditional independence between $Y_t^+$ and $Y_t^-$ for all $t=2,...,T$;
                thus, \(P(Y_t=y_t|Y_{t-1}=y_{t-1};\theta^+,\theta^-)=P(Y_t^+=y_t^+|Y_{t-1}=y_{t-1};\theta^+)P(Y_t^-=y_t^-|Y_{t-1}=y_{t-1};\theta^-)\)
        \end{itemize}
        Models which have a such form called the STERGM.
    \end{defn}
\end{frame}

\begin{frame}{BSTERGM: Bayesian STERGM}
    To convert the STERGM to the Bayesian setting, put priors $p(\theta^+),p(\theta^-)$ over $\theta^+,\theta^-$ and take the Bayes theorem as a inference rule.
    Then, the posterior of $\theta^+,\theta^-$ becomes
    \[P(\theta^+,\theta^-|y_t, y_{t-1}) = \frac{P(Y_t^+=y_t^+|y_{t-1},\theta^+) P(Y_t^-=y_t^-|y_{t-1},\theta^-)P(\theta^+),P(\theta^-)}{c(\theta^+,y_{t-1})c(\theta^-,y_{t-1})} \]
    where $y_t=y_t^+ - (y_{t-1} - y_t^-) = y_t^- + (y_t^+ - y_{t-1})$.

@@

\begin{frame}{Network Statistics for BSTERGM}
    As a usual, $s(y_{t}^.,y_{t-1})$ would be chosen by a difference of network statistics, $s'(y_t^.)-s'(y_{t-1})$.
    A common candidates of $s'$ are:
    \begin{itemize}
        \item the number of edges
        \item node degree distribution (each order)
        \item edgewise shared partner distribution (each order)
        \item dyadwise shared partner distribution (each order)
        \item k-star distribution
        \item triangle distribution
    \end{itemize}
    In fact, a bundle of all order of node degree distribution and all order of edgewise shared partner distribution form sufficient statistics of a graph.
    Also, we can set $s'$ by a function of these statistics.
\end{frame}

@@

    How do we find the posterior distribution of $\theta^+,\theta^-$?

    Note that an ordinary MCMC algorithm does not work in our situation, because the constant part remains when calculating the MCMC ratio.
    To solve this doubly-intractable constant problem, we should use more special MCMC technique, the exchange MCMC algorithm.
\end{frame}

\begin{frame}{Fitting the BSTERGM}
    Here is the algorithm of our main chain for the exchange algorithm.
    \begin{algo}[the main chain]
    Let $y_1,...,y_T$ be given. For $m=1,...,M$,
    \begin{enumerate}
        \item Propose candidates $\theta_*^+,\theta_*^-$ from $\epsilon(.|\theta_{m-1}^+,\theta_{m-1}^-)$.
        \item Select a lag $(t-1,t)$ randomly on $2 \leq t \leq T$.
        \item Generate an exchange graph $y_{ex} \in\mathcal{Y}|_t$ (with $y_{ex}^+, y_{ex}^-$) at the $\theta_*^+,\theta_*^-$.
        \item Calculate the exchange MCMC ratio $\pi$ at the lag,
            \[\pi = \frac{P(y_t^+|y_{t-1},\theta_*^+)P(y_t^-|y_{t-1},\theta_*^-)p(\theta_*^+,\theta_*^-)}
                {P(y_t^+|y_{t-1},\theta_{m-1}^+)P(y_t^-|y_{t-1},\theta_{m-1}^-)p(\theta_{m-1}^+,\theta_{m-1}^-)}
                \frac{P(y_{ex}^+|y_{t-1},\theta_{m-1}^+)P(y_{ex}^-|y_{t-1},\theta_{m-1}^-)}{P(y_{ex}^+|y_{t-1},\theta_*^+)P(y_{ex}^-|y_{t-1},\theta_*^-)}\]
        \item With probability $min(\pi,1)$, accept the proposal and put $(\theta_m^+,\theta_m^-) = (\theta_*^+,\theta_*^-)$.\\
            Otherwise, reject the proposal and put $(\theta_m^+,\theta_m^-) = (\theta_{m-1}^+,\theta_{m-1}^-)$.
    \end{enumerate}
    \end{algo}    
    Observe that, the $\pi$ has no normalizing constant terms because they are canceled out by added exchange terms.
    \(log\pi = (\theta_*^+-\theta_{m-1}^+)(s(y_{t}^+,y_{t-1})-s(y_{ex}^+,y_{t-1}))
    +(\theta_*^- -\theta_{m-1}^-)(s(y_{t}^-,y_{t-1})-s(y_{ex}^-,y_{t-1}))+log \frac{P(\theta_*^+,\theta_*^-)}{P(\theta_{m-1}^+,\theta_{m-1}^-)}\)
    \( = (\theta_*^+-\theta_{m-1}^+)(s'(y_{t}^+)-s'(y_{ex}^+))
    +(\theta_*^- -\theta_{m-1}^-)(s'(y_{t}^-)-s'(y_{ex}^-))+log \frac{P(\theta_*^+,\theta_*^-)}{P(\theta_{m-1}^+,\theta_{m-1}^-)}\)
\end{frame}

\begin{frame}{Fitting the BSTERGM}
    We need one more thing. To generate the exchange graph $y_{ex}$ for time $t$ (of the second part of main chain), 
    we should have a generative algorithm at a given parameter points.
    However, we still do not know the normalizing constant. Thus, I use one more MCMC chain.
    \begin{algo}[the auxiliary chain]
    Let $\theta^+,\theta^-,y_{t-1}$ be given. For $k=1,...,K$,
    \begin{enumerate}
        \item Select one edge randomly from the $y_{k-1}$, say, $y_{k-1;ij}$.
        \item Propose a new graph $y_*$ with switching the $y_{ij}$ value from $y_{k-1}$: 
        \\ If $y_{k-1;ij}=1$, then set $y_{*;ij}=0$(dissolution case.) Oterwise, If $y_{k-1;ij}=0$, then set $y_{*;ij}=1$(formation case.)
        \\ (If sample graphs are undirected, switch $y_{ji}$ simultaneously.)
        \item Take $\theta$ as $\theta^+$ or $\theta^-$ according to the case. Calculate the MCMC ratio $\phi$,
        \[\phi = \frac{P(Y_t=y_*|y_{t-1},\theta)}{P(Y_t=y_{k-1}|y_{t-1},\theta)}= \frac{exp(\theta^T s(y_*,y_{t-1}))}{exp(\theta^T s(y_{k-1},y_{t-1}))} = exp(\theta^T (s'(y*)-s'(y_{k-1})))\]
        \item With probability $min(\phi,1)$, accept the proposal and put $y_k=y_*$.\\
            Otherwise, reject the proposal and put $y_k=y_{k-1}$.
    \end{enumerate}
    \end{algo}
    After $K$ iteration, use the last network as the exchange sample in the second part of the main algorithm.
\end{frame}

@@

\section{Model inference}
\begin{frame}{Inference and Prediction}
    Since we have posterior samples by running the main chain as many as we want, we can do basic inference procedure.
    \begin{itemize}
        \item A outlining shape of the posterior of $\theta^+,\theta^-|y_1,y_2,...,y_T$ by histogram.
        \item An approximated summary statistics: mean, mode, variance, ...
        \item An approximated quantile and probability interval
    \end{itemize}

@@

\begin{frame}{Goodness of Fit}
    To evaluate the goodness of fit for the posterior $\theta^+,\theta^-$,
    we can use the auxiliary chain algorithm once again.
    \begin{algo}[GOF procedure]
        For t=2,...,T
        \\For s=1,...,S
        \begin{enumerate}
            \item sample $\theta_s^+,\theta_s^-$ from the estimate of posterior.
            \item simulate $y_s$ using the auxiliary chain under $y_{t-1}$.
            \item calculate $g(y_s)$, some higher degree statistics (eg. Node-degree dist \& Edgewise Shared Partner dist)
        \end{enumerate}
        Draw the box-plot of $g(y_s)$ and compare with $g(y_t)$.
    \end{algo}
\end{frame}

>>>>

\section{Model definition}
\begin{frame}{ERGM: Exponential family Random Graphs Models}
    \begin{defn}[ERGM: Exponential family Random Graphs Models]
        Let $\mathcal{Y} \subset \{0,1\}^{n^2}$ be a set of graphs with $n$ nodes and $Y$ be a random graph (on $\mathcal{Y}$.)

        Set a distribution on $\mathcal{Y}$ to
        \[P(Y=y;\theta) = \frac{exp(\theta^{T}s(y))}{c(\theta)}\]
        for some $\theta\in\mathbb{R}^p$,
        where $s(y)\in\mathbb{R}^p$ is a vector which is part of y's sufficient network statistics,
        \\ and $c(\theta)\in\mathbb{R}$ is a normalizing constant satisfying $c(\theta)=\sum_{y\in\mathcal{Y}}exp(\theta^{T}s(y))$.
        \\ Models which have a such form are called the ERGM.
    \end{defn}
\end{frame}

\begin{frame}{STERGM: Separable-Temporal Exponential family Random Graphs Models}
    \begin{defn}[STERGM: Separable-Temporal Exponential family Random Graphs Models]
        Let $Y_{t-1}=y_{t-1} \in \mathcal{Y}$ be given and $Y_t \in \mathcal{Y}$.
        Let $\mathcal{Y}^+ \subset \mathcal{Y}$ be a set of all graphs having equal or additional edges comparing to $y_{t-1}$. Likewise, define $\mathcal{Y}^-\subset\mathcal{Y}$ having equal or sparser edges.
        For $Y_t^+ \in \mathcal{Y}^+$, $Y_t^- \in \mathcal{Y}^-$, set

        Formation model:
        \[P(Y_t^+=y_t^+|Y_{t-1}=y_{t-1};\theta^+) = \frac{exp((\theta^+)^{T}s(y_t^+, y_{t-1}))}{c(\theta^+, y_{t-1})}\]
        Dissolution model:
        \[P(Y_t^-=y_t^-|Y_{t-1}=y_{t-1};\theta^-) = \frac{exp((\theta^-)^{T}s(y_t^-, y_{t-1}))}{c(\theta^-, y_{t-1})}\]
        for $\theta^+,\theta^-\in\mathbb{R}^p$, where $s(y_t^+, y_{t-1}), s(y_t^-, y_{t-1})\in\mathbb{R}^n$ are (functions of) sufficient network statistics,
        and normalizers $c(\theta^+, y_{t-1})=\sum_{y^+\in\mathcal{Y}^+}exp((\theta^+)^{T}s(y^+, y_{t-1})), c(\theta^-, y_{t-1})=\sum_{y^-\in\mathcal{Y}^-}exp((\theta^-)^{T}s(y^-, y_{t-1}))$.

        Then, defining operations $+,-$ on $\mathcal{Y}$ by the boolean algebra edgewisely, set $y_t$ to
        \[y_t=y_t^+ - (y_{t-1} - y_t^-) = y_t^- + (y_t^+ - y_{t-1})\]

        Additionally, assume that
        \begin{itemize}
            \item The first-order Markov assumption: $P(Y_2,...,Y_T|Y_1)=P(Y_2|Y_1)...P(Y_T|Y_{T-1})$
            \item The separability: the conditional independence between $Y_t^+$ and $Y_t^-$ for all $t=2,...,T$;
                thus, \(P(Y_t=y_t|Y_{t-1}=y_{t-1};\theta^+,\theta^-)=P(Y_t^+=y_t^+|Y_{t-1}=y_{t-1};\theta^+)P(Y_t^-=y_t^-|Y_{t-1}=y_{t-1};\theta^-)\)
        \end{itemize}
        Models which have a such form are called the STERGM.
    \end{defn}
\end{frame}

\begin{frame}{BSTERGM: Bayesian STERGM}
    To convert the STERGM to the Bayesian setting, put priors $p(\theta^+),p(\theta^-)$ over $\theta^+,\theta^-$ and take the Bayes rule as the inference rule.
    Then, the posterior of $\theta^+,\theta^-$ becomes
    \[P(\theta^+,\theta^-|y_t, y_{t-1}) = \frac{P(Y_t^+=y_t^+|y_{t-1},\theta^+) P(Y_t^-=y_t^-|y_{t-1},\theta^-)P(\theta^+),P(\theta^-)}{c(\theta^+,y_{t-1})c(\theta^-,y_{t-1})} \]
    where $y_t=y_t^+ - (y_{t-1} - y_t^-) = y_t^- + (y_t^+ - y_{t-1})$.

@@

\begin{frame}{Network Statistics for BSTERGM}
    As a usual, $s(y_{t}^.,y_{t-1})$ would be chosen by a difference of network statistics, $s'(y_t^.)-s'(y_{t-1})$.
    For example, some candidates of $s'$ are:
    \begin{itemize}
        \item the number of edges
        \item k-stars
        \item triangles
    \end{itemize}
    Note that we can set $s'$ by a function of sufficient network statistics.
\end{frame}

@@

    How do we find the posterior distribution of $\theta^+,\theta^-$?
\end{frame}

\begin{frame}{Fitting the BSTERGM}
    \begin{algo}[the main chain]
    Let $y_1,...,y_T$ be given. For $m=1,...,M$,
    \begin{enumerate}
        \item Propose candidates $\theta_*^+,\theta_*^-$ from $\epsilon(.|\theta_{m-1}^+,\theta_{m-1}^-)$.
        \item Select a lag $(t-1,t)$ randomly on $2 \leq t \leq T$.
        \item Generate an exchange graph $y_{ex} \in\mathcal{Y}$ (with $y_{ex}^+, y_{ex}^-$) at the $\theta_*^+,\theta_*^-$.
        \item Calculate the exchange MCMC ratio $\pi$ at the lag,
            \[\pi = \frac{P(y_t^+|y_{t-1},\theta_*^+)P(y_t^-|y_{t-1},\theta_*^-)p(\theta_*^+,\theta_*^-)}
                {P(y_t^+|y_{t-1},\theta_{m-1}^+)P(y_t^-|y_{t-1},\theta_{m-1}^-)p(\theta_{m-1}^+,\theta_{m-1}^-)}
                \frac{P(y_{ex}^+|y_{t-1},\theta_{m-1}^+)P(y_{ex}^-|y_{t-1},\theta_{m-1}^-)}{P(y_{ex}^+|y_{t-1},\theta_*^+)P(y_{ex}^-|y_{t-1},\theta_*^-)}\]
        \item With probability $min(\pi,1)$, accept the proposal and put $(\theta_m^+,\theta_m^-) = (\theta_*^+,\theta_*^-)$.\\
            Otherwise, reject the proposal and put $(\theta_m^+,\theta_m^-) = (\theta_{m-1}^+,\theta_{m-1}^-)$.
    \end{enumerate}
    \end{algo}    
    Observe that the $\pi$ has no normalizing constant terms because they are canceled out by added exchange terms.
    \(log\pi = (\theta_*^+-\theta_{m-1}^+)(s'(y_{t}^+)-s'(y_{ex}^+))
    +(\theta_*^- -\theta_{m-1}^-)(s'(y_{t}^-)-s'(y_{ex}^-))+log \frac{P(\theta_*^+,\theta_*^-)}{P(\theta_{m-1}^+,\theta_{m-1}^-)}\)

    Moreover, it is noteworthy that for the second step, an auxiliary MCMC chain which generate an exchange sample is used.
\end{frame}

% \begin{frame}{Fitting the BSTERGM}
%     We need one more thing. To generate the exchange graph $y_{ex}$ for time $t$ (of the second part of main chain), 
%     we should have a generative algorithm at a given parameter points.
%     However, we still do not know the normalizing constant. Thus, I use one more MCMC chain.
%     \begin{algo}[the auxiliary chain]
%     Let $\theta^+,\theta^-,y_{t-1}$ be given. For $k=1,...,K$,
%     \begin{enumerate}
%         \item Select one edge randomly from the $y_{k-1}$, say, $y_{k-1;ij}$.
%         \item Propose a new graph $y_*$ with switching the $y_{ij}$ value from $y_{k-1}$: 
%         \\ If $y_{k-1;ij}=1$, then set $y_{*;ij}=0$(dissolution case.) Oterwise, If $y_{k-1;ij}=0$, then set $y_{*;ij}=1$(formation case.)
%         \\ (If sample graphs are undirected, switch $y_{ji}$ simultaneously.)
%         \item Take $\theta$ as $\theta^+$ or $\theta^-$ according to the case. Calculate the MCMC ratio $\phi$,
%         \[\phi = \frac{P(Y_t=y_*|y_{t-1},\theta)}{P(Y_t=y_{k-1}|y_{t-1},\theta)}= \frac{exp(\theta^T s(y_*,y_{t-1}))}{exp(\theta^T s(y_{k-1},y_{t-1}))} = exp(\theta^T (s'(y*)-s'(y_{k-1})))\]
%         \item With probability $min(\phi,1)$, accept the proposal and put $y_k=y_*$.\\
%             Otherwise, reject the proposal and put $y_k=y_{k-1}$.
%     \end{enumerate}
%     \end{algo}
%     After $K$ iteration, use the last network as the exchange sample in the second part of the main algorithm.
% \end{frame}

@@

\section{Model inference}
\begin{frame}{Inference and Prediction}
    Since we have posterior samples by running the main chain as many as we want, we can follow the common inference procedure for the Bayes models.
    \begin{itemize}
        \item An outlining shape of the posterior of $\theta^+,\theta^-|y_1,y_2,...,y_T$ by histogram.
        \item An approximated summary statistics: mean, mode, variance, ...
        \item An approximated quantile and probability interval
    \end{itemize}

@@

\begin{frame}{Goodness of Fit}
    To evaluate the goodness of fit for the posterior of $\theta^+,\theta^-$,
    we can use the auxiliary chain algorithm once again.
    \begin{algo}[GOF procedure]
        For t=2,...,T
        \\For s=1,...,S
        \begin{enumerate}
            \item sample $\theta_s^+,\theta_s^-$ from the estimate of posterior.
            \item simulate $y_s$ using the auxiliary chain under $y_{t-1}$.
            \item calculate $g(y_s)$, some higher degree statistics (eg. Node-degree dist \& Edgewise Shared Partner dist)
        \end{enumerate}
        Draw a box-plot of $g(y_s)$s and compare with $g(y_t)$.
    \end{algo}
\end{frame}
